Analysis of gene regulatory properties underlying pleiotropy between complex traits

### 2022-10-19 presentation_gold2022_paris/presentation_gold2022_paris.tex
ENH

% Change size of footnotes
\renewcommand{\footnotesize}{\fontsize{5pt}{5pt}\selectfont}
\title{Analysis of gene regulatory properties underlying trait pleiotropy}
\subtitle{Meeting GOLD 2022}
\author{Aitor Gonz\'alez}
\institute{Aix Marseille Univ, INSERM, TAGC}
\date{Oct. 19, 2022}

@@

\begin{itemize}
\item Genetic alleles with higher frequency in individuals with a given disease
\item A very active field with many studies
\item Many genetic variants are involved in several phenotypes
\end{itemize}
%
\vfill
%
Limitations
%
\begin{itemize}
\item Correlation between variants
\item Which cell types are causal to the disease?
\item Over 90\% GWAS variants fall in non-coding regions
\end{itemize}
%
\vfill
%
These limitations can be partially addressed by looking at eQTLs in different tissues

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Colocalisation analysis of GWAS and eQTL variants}

@@

\begin{enumerate}
\item Large scale colocalization analysis between GWAS and eQTL variants
\item To identify pleiotropic genetic loci
\item To characterize the regulatory causes of trait pleiotropy
\end{enumerate}
\end{frame}

@@

GWAS/eQTL colocalization analysis allows to cluster diseases coherently

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Why are these variants pleiotropic? Do pleiotropic variants associate to more egenes?}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Do pleiotropic variants associate to more etissues?}

@@

This could explain why pleiotropic variants active in more etissues and associated to more egenes

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Conclusions}

Regulatory characteristics of pleiotropic variants
%
\begin{itemize}
\item Variant pleiotropy is correlated in regions
\item Enrichment of variants upstream (incl. 5'UTR) and downstream of genes (incl. 3'UTR)
\item Enrichment of transcription factor and CRM binding
\item Increased number of associated egenes and etissues
\end{itemize}
%
\vfill
%
Other outcomes of this study
%
\begin{itemize}
\item A large analysis of GWAS/eQTL colocalisations
\item A list of pleiotropic variants and regions
\end{itemize}

@@

\begin{itemize}
\item L\'eopoldine Lecerf (M1)
\item P Paul, P Rihet, M Michel, S Marquet, S Spicuglia
\end{itemize}
%
\vfill
%
Funding
%
\begin{itemize}
\item Institut Cancer et Immunologie (ICI)
\item Agence nationale de la recherche (ANR)
\item GOLD
\end{itemize}

>>>>

% Change size of footnotes
\renewcommand{\footnotesize}{\fontsize{5pt}{5pt}\selectfont}
\title{Analysis of gene regulatory properties underlying pleiotropy between complex traits}
\subtitle{Meeting GOLD 2022}
\author{Aitor Gonz\'alez}
\institute{Aix Marseille Univ, INSERM, TAGC}
\date{Oct. 19, 2022}

@@

\begin{itemize}
\item Genetic alleles with higher frequency in individuals with a given disease
\item A very active field with many studies
\item Many genetic variants are involved in several phenotypes
\end{itemize}
%
\vfill
%
Questions:
%
\begin{itemize}
\item Correlation between variants. Which is the causal variant?
\item Which cell type important for disease?
\item Over 90\% GWAS variants are non-coding. What is their role?
\end{itemize}
%
\vfill
%
Questions partially addressed by looking at eQTLs in different tissues

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Annotation of GWAS variants with gene and tissue targets using colocalization analysis}

@@

\begin{enumerate}
\item Large scale gene and tissue annotation of variants of complex traits using GWAS/eQTL colocalization analysis
\item To identify genetic variants leading to pleiotropy of complex traits
\item To characterize the gene regulatory mechanisms of trait pleiotropy
\end{enumerate}
\end{frame}

@@

\begin{itemize}
\item Disease clusters are coherent $\rightarrow$ Validates approche and data
\item Definition of disease classes
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{How are pleiotropic variants defined?}

\begin{itemize}
\item Manually annotate variants based on trait clustering and ICD in 129 classifications
\item Aggregate colocalized variants based on trait classes
\item Analysed variants in groupes based on the number of trait classes
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Do pleiotropic variants associate to more eQTL genes?}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Do pleiotropic variants associate to more eQTL tissues?}

@@

@@

\begin{itemize}
\item This could explain why pleiotropic variants active in more etissues and associated to more egenes
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Concluding remarks}

Outcomes interesting for the GOLD population genotypes
%
\begin{itemize}
\item A large set of causal GWAS variants annotated with target genes and tissues
\item A list of pleiotropic variants and regions
\end{itemize}
%
\vfill
%
Conclusion about regulatory characteristics of pleiotropic variants
%
\begin{itemize}
\item Variant pleiotropy is correlated in regions
\item Enrichment of variants upstream (incl. 5'UTR) and downstream of genes (incl. 3'UTR)
\item Enrichment of transcription factor and CRM binding
\item Increased number of associated egenes and etissues
\end{itemize}

@@

\begin{itemize}
\item L\'eopoldine Lecerf (M1)
\item P Paul, P Rihet, M Michel, S Marquet, S Spicuglia
\end{itemize}
%
\vfill
%
Funding
%
\begin{itemize}
\item Institut Cancer et Immunologie - Aix-Marseille Univ.
\item Agence nationale de la recherche (ANR)
\item GOLD
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{}
